---
title: "hMDM Cytokine Analysis Script"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Setup

```{r eval = FALSE}
rm(list = ls(all.names = TRUE)) # clears global environ.
```
```{r load-packages, include = FALSE}
# Load packages
library(tidyverse) # for data cleaning
library(pheatmap) # for making heatmap
library(viridis) # for heatmap color scheme
library(table1) # for making cytokine summary table
library(ggfortify) # for PCA
library(factoextra) # for PCA 
library(ggplot2) # for plotting
```
```{r eval = FALSE}
# Load packages
library(tidyverse) # for data cleaning
library(pheatmap) # for making heatmap
library(viridis) # for heatmap color scheme
library(table1) # for making cytokine summary table
library(ggfortify) # for PCA
library(factoextra) # for PCA 
library(ggplot2) # for plotting
```

### Import data

Set up data frame.
```{r}
MDMCytokineData <- read.csv("2021_05_20 MSD MDM Table.csv")
MDMCytokineData_df <- data.frame(MDMCytokineData, row.names = MDMCytokineData$Clean.Sample.ID)
MDMCytokineData_df <- MDMCytokineData_df[, 3:30]
head(MDMCytokineData_df)
```

Create summary data frame that contains the means of each mediator within each group.
```{r}
MDM_mean_df = MDMCytokineData_df %>% group_by(Polarization) %>% summarize(across(everything(), mean))
MDM_mean_df <- data.frame(MDM_mean_df[,-1], row.names = MDM_mean_df$Polarization)
head(MDM_mean_df)
```

### Summary Table of Concentrations in pg/mL

Create function so that values displayed in table are mean (SEM).
```{r}
my.render.cont.mean.sem <- function(x) {
  s <- stats.default(x)
  s$SEM <- with(s, SD/sqrt(N))
  with(stats.apply.rounding(s), c("",
                                  "Mean (SEM)"=sprintf("%s (%s)", MEAN, SEM)))
}
```

Use the output of this line of code to copy and paste into table1 function.
```{r}
paste0(" ", names(MDMCytokineData_df[2:ncol(MDMCytokineData_df)]), " ", collapse="+")
```

Make table. Row labels, including Greek symbols, as well as excess spacing, were fixed in Word. Significance markers were also added in Word using output from analysis in GraphPad Prism 9. 
```{r}
table1(~ Eotaxin + Eotaxin.3 + GM.CSF + IL.1a + IL.1B + IL.2 + IL.5 + IL.6 + IL.7 + IL.8 + IL.10 
       + IL.12p40 + IL.12p70 + IL.13 + IL.15 + IL.16 + IL.17 + IP.10 + MCP.1 + MCP.4 + MDC
       + MIP.1a + MIP.1B + TARC + TNF.a + TNF.B + VEGF | Polarization, 
       data = MDMCytokineData_df, 
       render.continuous = my.render.cont.mean.sem,
       render.missing = NULL,
       overall = NULL)
```


### Heatmap

Prepare data frame of cytokine means for pheatmap by formatting column names in the data frame and converting object to data matrix.
```{r}
# Delete periods from all of the columns
names(MDM_mean_df) <- gsub("\\.", "", names(MDM_mean_df))

# Turn data frame into matrix for pheatmap input
MDM_mean_dm <- data.matrix(MDM_mean_df)
```

Make heat map. Heat map was saved as PDF and Illustrator was used to add signficance stars to cells.
```{r}
pheatmap(t(MDM_mean_dm), 
         color = viridis(100), # sets color scheme
         angle_col = c("0"), # column labels horizontal
         cellwidth = 75, # sets dimensions of cells so that they don't change with viewer pane size
         cellheight = 10, # sets dimensions of cells so that they don't change with viewer pane size
         border_color = "black", # adds black border around cells
         treeheight_col = 10, # sets dims of trees so that they don't change with viewer pane size
         fontsize_row = 9, # sets dims of trees so that they don't change with viewer pane size
         scale = 'row', # scales data by row
         fontsize_col = 12, # sets font size for column labels
         cutree_rows = 5) # indicates how many clusters to show separated by spaces
```

### PCA 

Prepare data frame of cytokine data for PCA.
```{r}
# Remove periods from column names
names(MDMCytokineData_df) <- gsub("\\.", "", names(MDMCytokineData_df))

# Create a data frame with only mediator columns
MDMDataPCA <- MDMCytokineData_df[, 2:28]

# Log transform mediator data so that PCA plot isn't as bunched up.
# Pseudocount of 1 added because some values are zero and you can't take the log of 0.
MDMDataPCA_log <- log(MDMDataPCA+1)
```

Run PCA. Centering ensures components are only looking at variance within the dataset. Scaling brings all variables to the same magnitude so that high abundance variables do not influence results more than low abundance variables. 
```{r}
pca.res <- prcomp(MDMDataPCA_log, center = TRUE, scale = TRUE)
```

Plot PCA clustering results.
```{r}
# Set theme
theme_set(theme_bw())

# Cluster plot. Export as 3 x 4 PDF.
fviz_pca_ind(pca.res,
             label = "none",
             habillage = MDMCytokineData_df$Polarization, 
             palette = c("#0A0A0A", "#0253F5", "#02D40D"),
             addEllipses = TRUE) + 
  ggtitle("PCA - Clustering") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        panel.border = element_rect(fill = NA, color = "black", size = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = c(0.15, 0.25),
        legend.title = element_blank(),
        legend.background = element_rect(fill = "white", 
                                                     size = 0.3, 
                                                     linetype = "solid",
                                                     color = "black"),
        legend.text = element_text(size = 10)) +
  xlim(-10, 10)
```

Plot PCA contributions. To decide which variables to label in M1 direction, the top 5 contributors from axes 1 and 2 were chosen since so many labels overlapped.
```{r}
# Plot top contributing cytokines by PCA axis
fviz_contrib(pca.res, choice = "var", axes = 1, top = 10)
fviz_contrib(pca.res, choice = "var", axes = 2, top = 10)

# Vector of labels to include in the plot
labels <- c("TARC", "MCP4", "MDC", "IL7", "MIP1a", "TNFa", "IL8", "IL6", "IL15", "IL10", "Eotaxin3", "IL5")

# Final contributions plot
fviz_pca_var(pca.res, col.var = "contrib",
             repel = TRUE, col.circle = NA, labelsize = 4,
             select.var = list(name = labels)) +
  ggtitle("PCA - Contributions") +
  scale_color_viridis(begin = 0, end = 0.6) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        panel.border = element_rect(fill = NA, color = "black", size = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "right",
        legend.text = element_text(size = 10), 
        aspect.ratio=1/1.75) +
  labs(color = "Contribution") +
  ylim(-1.25, 1.25) +
  xlim(-1.25, 1.25)
```